---
title: "COVID_19"
author: "Milad"
date: "2023-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question Statement

In this R markdown file, we want to do some basic explanatory analysis of the John Hopkins University COVID_19 dataset. We try to visualize the growth of cases across United States and inter_countries comparison of cases. We'll provide one modeling of the data and will try to identify any potential biases at the end.


## Data Source and its description

This is the data repository for the 2019 Novel Coronavirus Visual Dashboard operated by the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Please note that On March 10, 2023, the Johns Hopkins Coronavirus Resource Center ceased its collecting and reporting of global COVID-19 data. For updated cases, deaths, and vaccine data please visit the following sources:

- Global: World Health Organization (WHO)
- U.S.: U.S. Centers for Disease Control and Prevention (CDC) 

This folder contains daily time series summary tables, including confirmed, deaths and recovered. All data is read in from the daily case report.

Two time series tables are for the US confirmed cases and deaths, reported at the county level. They are named time_series_covid19_confirmed_US.csv, time_series_covid19_deaths_US.csv, respectively.

Three time series tables are for the global confirmed cases, recovered cases and deaths. Australia, Canada and China are reported at the province/state level. Dependencies of the Netherlands, the UK, France and Denmark are listed under the province/state level. The US and other countries are at the country level. The tables are renamed time_series_covid19_confirmed_global.csv and time_series_covid19_deaths_global.csv, and time_series_covid19_recovered_global.csv, respectively.

## Importing Libraries

```{r Libraries , warning=FALSE, message=FALSE}
if (!require(tidyverse)) {
  install.packages("tidyverse" , repos = "http://cran.us.r-project.org")
}
library(tidyverse)

if (!require(lubridate)) {
  install.packages("lubridate" , repos = "http://cran.us.r-project.org")
}
library(lubridate)
```

## Importing Data


```{r Import}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv","time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)
urls
```

## Defining different datasets

```{r defining_datasets ,warning=FALSE, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```

## Tidying up the datasets

As stated in the lectures, we would like to tidy those datasets and put each variable (date, cases, deaths) in their own column. We will also get rid of the longitude and latitude attributes since we don't need themfor our analysis. We'll also rename Region and State. Note that by pivoting other columns, we are having each separate day as a row.

```{r cleaning}
global_cases <- global_cases %>%
              pivot_longer(cols= -c('Province/State','Country/Region','Lat','Long'),
                           names_to = "date",
                           values_to = "cases") %>%
              select(-c(Lat,Long))
global_cases

global_deaths <- global_deaths %>%
              pivot_longer(cols= -c('Province/State','Country/Region','Lat','Long'),
                           names_to = "date",
                           values_to = "deaths") %>%
              select(-c(Lat,Long))
global_deaths

```

## Joining the global datasets

```{r joining_global_datasets}

global <- global_cases %>%
          full_join(global_deaths) %>%
          rename(Country_Region = 'Country/Region',
                    Province_State = 'Province/State') %>%
          mutate(date = mdy(date))

global
summary(global)

```

```{r further_cleaning_on_global_dataset}

global <- global %>% filter(cases > 0)
summary(global)
```

## Cleaning the US datasets

```{r cleaning_US_datasets}
US_cases %>% 
        pivot_longer(cols= -(UID:Combined_Key),
                     names_to = "date",
                     values_to="cases") %>%
        select(Admin2:cases) %>%
        mutate(date=mdy(date)) %>%
        select(-c(Lat, Long_))

US_deaths %>% 
        pivot_longer(cols= -(UID:Population),
                     names_to = "date",
                     values_to="deaths") %>%
        select(Admin2:deaths) %>%
        mutate(date=mdy(date)) %>%
        select(-c(Lat, Long_))
US_cases
US_deaths

```

## Joining the US datasets

```{r joining_us_datasets , message=FALSE}
US <- US_cases %>%
      full_join(US_deaths)
US
```

## Making the Global and US datasets compatible and adding the population

```{r datasets_compatibility}
global <- global %>% 
          unite("Combined_Key",
                c(Province_State, Country_Region),
                sep = ", ",
                na.rm = TRUE,
                remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
        select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>% 
          left_join(uid, by=c("Province_State", "Country_Region")) %>%
          select(-c(UID, FIPS)) %>%
          select(Province_State, Country_Region , date, 
                 cases, deaths, Population,Combined_Key)

global

```

## Visualizations

```{r visual_1}
selected_country <- "US"

country_data <- global %>%
  filter(Country_Region == selected_country)

# Plot number of cases and deaths over time
country_data %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = cases), color = "blue") +
  geom_line(aes(y = deaths), color = "red") +
  scale_y_continuous(sec.axis = sec_axis(~. , name = "Deaths")) +
  labs(x = "Date", y = "Cases", title = paste("COVID-19 in", selected_country))

# Summarize data by country
country_totals <- global %>%
  group_by(Country_Region) %>%
  summarise(total_cases = max(cases, na.rm = TRUE))

# Plot total cases by country
country_totals %>%
  arrange(desc(total_cases)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(Country_Region, -total_cases), y = total_cases)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Country", y = "Total Cases", title = "Total COVID-19 Cases by Country")

```


## Modeling the data

```{r model}
# Create a new variable for the number of days since the first case
country_data <- country_data %>%
  mutate(days_since_first_case = as.numeric(difftime(date, min(date, na.rm = TRUE), units = "days")))

# Fit a linear regression model
fit <- lm(cases ~ days_since_first_case, data = country_data)

# Print the summary of the model
summary(fit)
```

## Conclusion

By just looking at growth over time for US COVID cases, other than a general increase as a trend which is expected, we also see a rather steep increase at early 2022(Jan & Feb) . 
By comparing total COVID cases of different countries, we can see that US stands at a significantly higher spot than other countries. 

## Identifying potential sources of bias

- Testing Rate Bias: The number of confirmed cases in a country heavily depends on the testing rate. Countries testing more people will likely report more cases, which doesn't necessarily mean they're the most affected.

- Reporting Bias: Different countries have different standards and capacities for reporting COVID-19 cases and deaths. Some countries might under-report due to a lack of resources, while others might over-report due to different definitions of what constitutes a COVID-19 case or death.

- Population Size Bias: Larger countries are likely to have more cases simply due to their larger population sizes. To compare countries more fairly, you should look at cases per capita.

- Time Series Bias: The spread of COVID-19 is not linear. The virus often spreads exponentially, so the use of a linear regression model could be misleading.

- Selection Bias: Focusing on one country (in first visual) and the top 10 countries (in second visual) only gives a limited perspective. It's crucial to consider the global context and other countries' situations.





