---
title: "COVID_19"
author: "Milad"
date: "2023-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing Libraries

```{r Libraries , warning=FALSE, message=FALSE}
if (!require(tidyverse)) {
  install.packages("tidyverse" , repos = "http://cran.us.r-project.org")
}
library(tidyverse)

if (!require(lubridate)) {
  install.packages("lubridate" , repos = "http://cran.us.r-project.org")
}
library(lubridate)
```

## Importing Data


```{r Import}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv","time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)
urls
```

## Defining different datasets

```{r defining_datasets ,warning=FALSE, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```

## Tidying up the datasets

As stated in the lectures, we would like to tidy those datasets and put each variable (date, cases, deaths) in their own column. We will also get rid of the longitude and latitude attributes since we don't need themfor our analysis. We'll also rename Region and State. Note that by pivoting other columns, we are having each separate day as a row.

```{r cleaning}
global_cases <- global_cases %>%
              pivot_longer(cols= -c('Province/State','Country/Region','Lat','Long'),
                           names_to = "date",
                           values_to = "cases") %>%
              select(-c(Lat,Long))
global_cases

global_deaths <- global_deaths %>%
              pivot_longer(cols= -c('Province/State','Country/Region','Lat','Long'),
                           names_to = "date",
                           values_to = "deaths") %>%
              select(-c(Lat,Long))
global_deaths

```

## Joining the global datasets

```{r joining_global_datasets}

global <- global_cases %>%
          full_join(global_deaths) %>%
          rename(Country_Region = 'Country/Region',
                    Province_State = 'Province/State') %>%
          mutate(date = mdy(date))

global
summary(global)

```

```{r further_cleaning_on_global_dataset}

global <- global %>% filter(cases > 0)
summary(global)
```

## Cleaning the US datasets

```{r cleaning_US_datasets}
US_cases %>% 
        pivot_longer(cols= -(UID:Combined_Key),
                     names_to = "date",
                     values_to="cases") %>%
        select(Admin2:cases) %>%
        mutate(date=mdy(date)) %>%
        select(-c(Lat, Long_))

US_deaths %>% 
        pivot_longer(cols= -(UID:Population),
                     names_to = "date",
                     values_to="deaths") %>%
        select(Admin2:deaths) %>%
        mutate(date=mdy(date)) %>%
        select(-c(Lat, Long_))
US_cases
US_deaths

```

## Joining the US datasets

```{r joining_us_datasets}
US <- US_cases %>%
      full_join(US_deaths)
US
```

## Making the Global and US datasets compatible and adding the population

```{r datasets_compatibility}
global <- global %>% 
          unite("Combined_Key",
                c(Province_State, Country_Region),
                sep = ", ",
                na.rm = TRUE,
                remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
        select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>% 
          left_join(uid, by=c("Province_State", "Country_Region")) %>%
          select(-c(UID, FIPS)) %>%
          select(Province_State, Country_Region , date, 
                 cases, deaths, Population,Combined_Key)

global

```

## Visualizations

```{r visual_1}
selected_country <- "US"

country_data <- global %>%
  filter(Country_Region == selected_country)

# Plot number of cases and deaths over time
country_data %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = cases), color = "blue") +
  geom_line(aes(y = deaths), color = "red") +
  scale_y_continuous(sec.axis = sec_axis(~. , name = "Deaths")) +
  labs(x = "Date", y = "Cases", title = paste("COVID-19 in", selected_country))

# Summarize data by country
country_totals <- global %>%
  group_by(Country_Region) %>%
  summarise(total_cases = max(cases, na.rm = TRUE))

# Plot total cases by country
country_totals %>%
  arrange(desc(total_cases)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(Country_Region, -total_cases), y = total_cases)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(x = "Country", y = "Total Cases", title = "Total COVID-19 Cases by Country")

```


## Modeling the data

```{r model}
# Create a new variable for the number of days since the first case
country_data <- country_data %>%
  mutate(days_since_first_case = as.numeric(difftime(date, min(date, na.rm = TRUE), units = "days")))

# Fit a linear regression model
fit <- lm(cases ~ days_since_first_case, data = country_data)

# Print the summary of the model
summary(fit)
```





